Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks

### 2024-12-24 _manuscript/_tex/index.tex
section 3

In data analysis, researchers often rely on their prior knowledge or
domain expertise to form expectations and assess whether results align
with them. When results deviate from these expectations, experienced
researchers check assumptions made about the data and during the
analysis, but these assumptions checked are often difficult to trace and
are seldom documented in publications. While publication of analysis
code and data is now a common a requirement for the sake of
reproducibility \citep{peng2011reproducible}, the published code alone
is often insufficient for understanding the thought process behind an
analysis. Code corresponding to published analyses often reflect the
final decisions made about analysis and do not reveal the decisions or
assumptions made about the data during the analysis. Thus, a reader
looking at published code can often be left with many questions about
why certain choices were made.

@@

Expectations represent our understanding of certain aspects of the
analysis and the data, independent of the results of the analysis
itself. When observed outcomes deviate from these expectations, analysts
often revisit the analysis process to identify potential issues, refine
methods, or revise assumptions. Experienced analysts can typically
identify issues quickly and correct them on the spot, but they often do
so without discussing the underlying reasoning, limiting transparency
and making it harder for less experienced researchers to learn and
master these skills.

Here, we introduce the concept of \textbf{analysis validation checks},
which frame these expectations or assumptions as explicit checks that
return a TRUE or FALSE result given the data analyzed at hand. Inspired
by the concept of data validation checks \citep{validate}, which are
designed to ensure that datasets meet expected formats and quality,
analysis validation checks reverse the approach: they validate the
assumptions about the data necessary for the analysis to produce the
\emph{expected results}, as defined by the analyst. The focus on
expected results allows the concept of analysis validation checks to
encompass a broad range of checks, such as data quality (i.e.~missing
data, how the data are structured), data distribution and outliers,
bivariate and multivariate relationships between variables, and other
contextual information.

@@

In addition to the above benefits, the development and publication of
analysis checks has the potential to help students, inexperienced
analysts, and junior researchers develop the skills needed to diagnose
unexpected analysis results for a given type of data because the
assumptions made about the data are made transparent. The analysis
checks can serve as a basis for new analysts to have conversations about
the data they are analyzing and to develop a better understanding of the
potential data generation process.

@@

Consider a 30-day step count experiment in a public health setting.
Subjects are instructed to walk at least 8,000 steps each day, with an
expected average of 9,000 steps, tracked by a step counter app. After 30
days, we review the data and examine the number of steps taken each day.
With data of this nature, we may expect there to be occasional ``low''
days due to factors such as forgetting to wear their watch or
unfavorable weather conditions limiting outdoor activities. We may also
expect ``high'' days recorded after an outdoor hike or intense workout.
Given the requirements of the study, we expect that for a given subject
the average step count would be between \([8500, 9500]\).

To diagnose potential reasons why this outcome expectation might fail,
we can establish a few analysis validation checks in anticipation of
seeing the data. For example, we can check the quantile of the step
count, if more than a third of the days fall below 8,000, or more than a
third exceed 10,000 steps, this could indicate an excess of low-count or
high-count days. Similarly, we may may expect the standard deviation of
the step count not to be overly large. These considerations yield the
following three analysis validation checks that \emph{fail} when:

\begin{itemize}
\tightlist
\item
  check 1: the 60\% quantile of the observed step counts is greater than
  10,000
\item
  check 2: the 40\% quantile of the observed step counts is less than
  8,000, and
\item
  check 3: the standard deviation of the observed step counts exceeds
  2,500.
\end{itemize}

The cutoff values chosen for these checks would presumably be chosen
based on prior experience with these kinds of data, but could also be
optimized using the method presented in the next section.

@@

\caption{\label{fig-step-count}Beeswarm plot of average step counts
across 300 simulated 30-day periods. Each point represents the average
step count from one simulation. While similar to a boxplot or violin
plot, the beeswarm plot also displays the distribution of each
individual data point. The orange points indicate instances where the
average step count fails outside the {[}8,500, 9,500{]} interval,
representing an unexpected outcome in this scenario.}

@@

To simulate this data, three normal distributions are used for the daily
step counts: \(\mathcal{N}(6000, 200)\) for low days,
\(\mathcal{N}(12000, 200)\) for high days, and
\(\mathcal{N}(9000, 300)\) for typical days. The number of low and high
days can be simulated from a Poisson distribution with \(\lambda = 8\).
Figure~\ref{fig-step-count} displays average step count across 300
simulated 30-day periods.

>>>>

In data analysis, analysts often rely on their prior knowledge or domain
expertise to form expectations and assess whether results align with
them. When results deviate from these expectations, experienced analysts
check assumptions made about the data and during the analysis, but often
do so without discussing the underlying reasoning in the publications.
While publication of analysis code and data is now a common a
requirement for the sake of reproducibility
\citep{peng2011reproducible}, the published code alone is often
insufficient for understanding the thought process behind an analysis.
Code corresponding to published analyses often reflect the final
decisions made about analysis and do not reveal the decisions or
assumptions made about the data during the analysis. Thus, a reader
looking at published code can often be left with many questions about
why certain choices were made.

@@

Analysis validation checks are assumptions made by the analysts during
the analysis, framed as explicit checks that return a binary TRUE or
FALSE value based on the data. Inspired by the concept of data
validation checks \citep{validate}, which are designed to ensure that
datasets meet expected quality before the analysis begins, analysis
validation checks reverse the approach: they validate the assumptions
about the data necessary for the analysis to produce the \emph{expected
results}, as defined by the analyst. The focus on expected results
allows analysis validation checks to encompass a wide range of checks,
such as data quality, variable distributions and outliers, bivariate and
multivariate relationships, and contextual information relevant to the
analysis.

@@

In addition to the above benefits, the development and publication of
analysis checks have the potential to help students, inexperienced
analysts, and junior researchers develop the skills needed to diagnose
unexpected analysis results for a given type of data because the
assumptions made about the data are made transparent. The analysis
checks can serve as a basis for new analysts to have conversations about
the data they are analyzing and to develop a better understanding of the
potential data generation process.

@@

Consider a 30-day step count experiment in a public health setting.
Subjects are instructed to walk at least 8,000 steps each day, with an
expected average of 9,000 steps, tracked by a step counter app. With
data of this nature, we may expect there to be occasional ``low'' days
due to factors such as unfavorable weather conditions limiting outdoor
activities. We may also expect ``high'' days recorded after an outdoor
activities or intense workout. Given the requirements of the study, we
form our expectation as follows:

\begin{quote}
Expectation: The average step count of a given subject is between
\([8,500, 9,500]\)
\end{quote}

To diagnose potential reasons why this expectation might fail, we can
establish a few analysis validation checks in anticipation of seeing the
data. For example, we can check the quantile of the step count, if more
than a third of the days fall below 8,000, or more than a third exceed
10,000 steps, this could indicate an excess of low-count or high-count
days. Similarly, we may may expect the standard deviation of the step
count not to be overly large. These considerations yield the following
three analysis validation checks that \emph{fail} when:

\begin{itemize}
\tightlist
\item
  Check 1: the 60\% quantile of the observed step counts is greater than
  10,000
\item
  Check 2: the 40\% quantile of the observed step counts is less than
  8,000, and
\item
  Check 3: the standard deviation of the observed step counts exceeds
  2,500.
\end{itemize}

The cutoff values for these checks would presumably be chosen based on
prior experience with these kinds of data, but could also be optimized
using the method presented in the next section.

To simulate this data, three normal distributions are used for the daily
step counts: \(\mathcal{N}(6,000, 200)\) for low days,
\(\mathcal{N}(12,000, 200)\) for high days, and
\(\mathcal{N}(9,000, 300)\) for typical days. The number of low and high
days can be simulated from a Poisson distribution with \(\lambda = 8\).
Figure~\ref{fig-step-count} displays average step count across 300
simulated 30-day periods.

@@

\caption{\label{fig-step-count}Beeswarm plot of average step counts
across 300 simulated 30-day periods. Each point represents the average
step count from one simulation. Similar to a boxplot or violin plot, the
beeswarm plot also displays the distribution of each individual data
point. The orange points indicate instances where the average step count
fails outside the {[}8,500, 9,500{]} interval, representing an
unexpected outcome in this scenario.}

@@

### 2024-12-24 index.qmd
section 3

In data analysis, researchers often rely on their prior knowledge or domain expertise to form expectations and assess whether results align with them. When results deviate from these expectations, experienced researchers check assumptions made about the data and during the analysis, but these assumptions checked are often difficult to trace and are seldom documented in publications. While publication of analysis code and data is now a common a requirement for the sake of reproducibility [@peng2011reproducible], the published code alone is often insufficient for understanding the thought process behind an analysis. Code corresponding to published analyses often reflect the final decisions made about analysis and do not reveal the decisions or assumptions made about the data during the analysis. Thus, a reader looking at published code can often be left with many questions about why certain choices were made.

@@

Expectations represent our understanding of certain aspects of the analysis and the data, independent of the results of the analysis itself. When observed outcomes deviate from these expectations, analysts often revisit the analysis process to identify potential issues, refine methods, or revise assumptions. Experienced analysts can typically identify issues quickly and correct them on the spot, but they often do so without discussing the underlying reasoning, limiting transparency and making it harder for less experienced researchers to learn and master these skills.

Here, we introduce the concept of **analysis validation checks**, which frame these expectations or assumptions as explicit checks that return a TRUE or FALSE result given the data analyzed at hand. Inspired by the concept of data validation checks [@validate], which are designed to ensure that datasets meet expected formats and quality, analysis validation checks reverse the approach: they validate the assumptions about the data necessary for the analysis to produce the *expected results*, as defined by the analyst. The focus on expected results allows the concept of analysis validation checks to encompass a broad range of checks, such as data quality (i.e. missing data, how the data are structured), data distribution and outliers, bivariate and multivariate relationships between variables, and other contextual information.

@@

Consider a 30-day step count experiment in a public health setting. Subjects are instructed to walk at least 8,000 steps each day, with an expected average of 9,000 steps, tracked by a step counter app. After 30 days, we review the data and examine the number of steps taken each day. With data of this nature, we may expect there to be occasional "low" days due to factors such as forgetting to wear their watch or unfavorable weather conditions limiting outdoor activities.  We may also expect "high" days recorded after an outdoor hike or intense workout. Given the requirements of the study, we expect that for a given subject 
the average step count would be between $[8500, 9500]$.

To diagnose potential reasons why this outcome expectation might fail, we can establish a few analysis validation checks in anticipation of seeing the data. For example, we can check the quantile of the step count, if more than a third of the days fall below 8,000, or more than a third exceed 10,000 steps, this could indicate an excess of low-count or high-count days. Similarly, we may may expect the standard deviation of the step count not to be overly large. These considerations yield the following three analysis validation checks that *fail* when:

-   check 1: the 60% quantile of the observed step counts is greater than 10,000
-   check 2: the 40% quantile of the observed step counts is less than 8,000, and
-   check 3: the standard deviation of the observed step counts exceeds 2,500.

The cutoff values chosen for these checks would presumably be chosen based on prior experience with these kinds of data, but could also be optimized using the method presented in the next section.

```{r}
#| fig-height: 3
#| label: fig-step-count
#| fig-cap: Beeswarm plot of average step counts across 300 simulated 30-day periods. Each point represents the average step count from one simulation. While similar to a boxplot or violin plot, the beeswarm plot also displays the distribution of each individual data point. The orange points indicate instances where the average step count fails outside the [8,500, 9,500] interval, representing an unexpected outcome in this scenario. 
set.seed(1234)
step_count <- tibble(id = 1:300) |> 
  rowwise() |> 
  mutate(
    small_n = rpois(1, 8),
    large_n = rpois(1, 8),
    norm_sd = rlogis(1, 800, scale = 100),
    norm_n = 30 - small_n - large_n,
    step = list(round(c(rnorm(small_n, 6000, 200), 
                        rnorm(large_n, 12000, 200), 
                        rnorm(norm_n, 9000, 300)
    ), 0)),
    mean = mean(step, na.rm = TRUE),
    unexpect = ifelse(between(mean, 8500, 9500), 0, 1),
    `q(step, 0.6) > 10000` = ifelse(quantile(step, 0.6, na.rm = TRUE) > 10000, 1, 0),
    `q(step, 0.4) < 8000` = ifelse(quantile(step, 0.4, na.rm = TRUE) < 8000, 1, 0),
    `sd(step) > 2500` = ifelse(sd(step) > 2500, 1, 0),
  ) |>
  ungroup()

@@

To simulate this data, three normal distributions are used for the daily step counts: $\mathcal{N}(6000, 200)$ for low days, $\mathcal{N}(12000, 200)$ for high days, and $\mathcal{N}(9000, 300)$ for typical days. The number of low and high days can be simulated from a Poisson distribution with $\lambda = 8$. @fig-step-count displays average step count across 300 simulated 30-day periods.

>>>>

In data analysis, analysts often rely on their prior knowledge or domain expertise to form expectations and assess whether results align with them. When results deviate from these expectations, experienced analysts check assumptions made about the data and during the analysis, but often do so without discussing the underlying reasoning in the publications. While publication of analysis code and data is now a common a requirement for the sake of reproducibility [@peng2011reproducible], the published code alone is often insufficient for understanding the thought process behind an analysis. Code corresponding to published analyses often reflect the final decisions made about analysis and do not reveal the decisions or assumptions made about the data during the analysis. Thus, a reader looking at published code can often be left with many questions about why certain choices were made.

@@

<!-- Expectations represent our understanding of certain aspects of the analysis and the data, independent of the results of the analysis itself. When observed outcomes deviate from these expectations, analysts often revisit the analysis process to identify potential issues, refine methods, or revise assumptions. Experienced analysts can typically identify issues quickly and correct them on the spot, but they often do so without discussing the underlying reasoning, limiting transparency and making it harder for less experienced researchers to learn and master these skills.  -->

Analysis validation checks are assumptions made by the analysts during the analysis, framed as explicit checks that return a binary TRUE or FALSE value based on the data. Inspired by the concept of data validation checks [@validate], which are designed to ensure that datasets meet expected quality before the analysis begins, analysis validation checks reverse the approach: they validate the assumptions about the data necessary for the analysis to produce the *expected results*, as defined by the analyst. The focus on expected results allows analysis validation checks to encompass a wide range of checks, such as data quality, variable distributions and outliers, bivariate and multivariate relationships, and contextual information relevant to the analysis.

@@

Consider a 30-day step count experiment in a public health setting. Subjects are instructed to walk at least 8,000 steps each day, with an expected average of 9,000 steps, tracked by a step counter app. With data of this nature, we may expect there to be occasional "low" days due to factors such as unfavorable weather conditions limiting outdoor activities.  We may also expect "high" days recorded after an outdoor activities or intense workout. Given the requirements of the study, we form our expectation as follows:

  > Expectation: The average step count of a given subject is between $[8,500, 9,500]$

To diagnose potential reasons why this expectation might fail, we can establish a few analysis validation checks in anticipation of seeing the data. For example, we can check the quantile of the step count, if more than a third of the days fall below 8,000, or more than a third exceed 10,000 steps, this could indicate an excess of low-count or high-count days. Similarly, we may may expect the standard deviation of the step count not to be overly large. These considerations yield the following three analysis validation checks that *fail* when:

-   Check 1: the 60% quantile of the observed step counts is greater than 10,000
-   Check 2: the 40% quantile of the observed step counts is less than 8,000, and
-   Check 3: the standard deviation of the observed step counts exceeds 2,500.

The cutoff values for these checks would presumably be chosen based on prior experience with these kinds of data, but could also be optimized using the method presented in the next section.

To simulate this data, three normal distributions are used for the daily step counts: $\mathcal{N}(6,000, 200)$ for low days, $\mathcal{N}(12,000, 200)$ for high days, and $\mathcal{N}(9,000, 300)$ for typical days. The number of low and high days can be simulated from a Poisson distribution with $\lambda = 8$. @fig-step-count displays average step count across 300 simulated 30-day periods.

```{r}
#| fig-height: 3
#| label: fig-step-count
#| fig-cap: Beeswarm plot of average step counts across 300 simulated 30-day periods. Each point represents the average step count from one simulation. Similar to a boxplot or violin plot, the beeswarm plot also displays the distribution of each individual data point. The orange points indicate instances where the average step count fails outside the [8,500, 9,500] interval, representing an unexpected outcome in this scenario. 
set.seed(1234)
step_count <- tibble(id = 1:300) |> 
  rowwise() |> 
  mutate(
    small_n = rpois(1, 8),
    large_n = rpois(1, 8),
    norm_sd = rlogis(1, 800, scale = 100),
    norm_n = 30 - small_n - large_n,
    step = list(round(c(rnorm(small_n, 6000, 200), 
                        rnorm(large_n, 12000, 200), 
                        rnorm(norm_n, 9000, 300)
    ), 0)),
    mean = mean(step, na.rm = TRUE),
    unexpect = ifelse(between(mean, 8500, 9500), 0, 1),
    `q(step, 0.6) > 10000` = ifelse(quantile(step, 0.6, na.rm = TRUE) > 10000, 1, 0),
    `q(step, 0.4) < 8000` = ifelse(quantile(step, 0.4, na.rm = TRUE) < 8000, 1, 0),
    `sd(step) > 2500` = ifelse(sd(step) > 2500, 1, 0),
  ) |>
  ungroup()

@@
